\pkg{plotROC}: A Better Tool for Plotting ROC Curves

### 2015-03-13 Paper/sachs2015plotROC.Rmd
response to first editorial review

---
title: '\pkg{plotROC}: A Better Tool for Plotting ROC Curves'
shorttitle: '\pkg{plotROC}: A Better Tool for Plotting ROC Curves'
plaintitle: 'plotROC: A Better Tool for Plotting ROC Curves'
author:
- name: Michael C. Sachs
  affil: Biometric Research Branch, Division of Cancer Treatment and Diagnosis, National Cancer Institute
address:
- Michael C. Sachs
- 9609 Medical Center Drive, MSC 9735
- Bethesda, MD 20892
- "telephone: 240-276-7888"
email: michael.sachs@nih.gov
abstract: 'Plots of the receiver operating characteristic (ROC) curve are ubiquitous in medical research. Designed to simultaneously display the operating characteristics at every possible value of a continuous diagnostic test, ROC curves are used in oncology to evaluate screening, diagnostic, prognostic and predictive biomarkers. I reviewed a sample of ROC curve plots from the major oncology journals in order to assess current trends in usage and design elements. My review suggests that ROC curve plots are often ineffective as statistical charts and that poor design obscures the relevant information the chart is intended to display. I describe my new \proglang{R} package that was created to address the shortcomings of existing tools. The package has functions to create informative ROC curve plots, with sensible defaults, for use in print or as an interactive web-based plot. A web application was developed to reach a broader audience of scientists who do not use R.'
keywords: ROC curves; graphics; interactive; plots
bibliography: plotroc
...

```{r setup, echo = FALSE}
library(knitr)
library(stringr)
knit_hooks$set(plot = function(x, options) {
  if ('tikz' %in% options$dev && !options$external) {
    hook_plot_tex(x, options)
  } else hook_plot_md(x, options)
})
opts_chunk$set(dev="tikz", comment = NA)
knit_hooks$set(source = function(x, options){
  if (!is.null(options$verbatim) && options$verbatim){
    opts = gsub(",\\s*verbatim\\s*=\\s*TRUE\\s*", "", options$params.src)
    opts = gsub(",\\s*eval\\s*=\\s*FALSE\\s*", "", opts)
    bef = sprintf('\n\n    ```{r %s}\n', opts, "\n")
    stringr::str_c(
      bef, 
      knitr:::indent_block(paste(x, collapse = '\n'), "    "), 
      "\n    ```\n"
    )
  } else {
    stringr::str_c("\n\n```", tolower(options$engine), "\n", 
      paste(x, collapse = '\n'), "\n```\n\n"
    )
  }
})
```

@@

## About ROC Curves

@@

## Design Considerations

@@

## Existing Plotting Software

The ROC curve plot is, at the most basic level, a line graph. Therefore, once the appropriate statistics are estimated, existing plotting functions can be used to create an ROC curve plot. Viewers can identify ROC plots through context, by observing the shape of the line, and through the addition of axis labels, titles, legends, and so on. In my review of the oncology literature, I observed plots with the distinctive characteristics of the plotting functions from Microsoft Office, \proglang{SAS}, \proglang{SPSS}, and the base \proglang{R} plotting functions [@arr].

There are several \proglang{R} packages related to ROC curve estimation that contain dedicated plotting functions. The \pkg{ROCR} package [@rocr] plots the FPF _versus_ TPF, as usual, and then takes the interesting approach of encoding the cutoff values as a separate color scale along the ROC curve itself. A legend for the color scale is placed along the vertical axis on the right of the plotting region. The \pkg{pROC} package [@pROC] is mainly focused on estimating confidence intervals and regions for restricted ranges of the ROC curve. The plotting methods therein use the base \proglang{R} plotting functions to create nice displays of the curves along with shaded confidence regions. My \pkg{plotROC} package uses the \pkg{ggplot2} [@ggplot2] plotting library to create clear, informative ROC plots, with interactive features for use on the web, and sensible defaults for use in print.

@@

Each image consisted of a single ROC curve plot or a panel of multiple plots. Each plot was inspected manually for the following design features: the number of curves displayed, the type of axis labels (sensitivity/ 1 - specificity or true/false positive fractions), presence or absence of grid lines, presence or absence of a diagonal guide line, whether any cutpoints were indicated, the type of curve label (legend or direct label), and presence of other textual annotations such as the AUC. The numerical results of the survey are summarized in table \ref{table1}.

@@

    if(tab) groupsa <- paste0("\t", groups) else groupsa <- groups
    blah <- data.frame(lab = groupsa, `percent (count)`  = NA, check.names = FALSE, stringsAsFactors = FALSE)
    N <- sum(!is.na(var))
    for(k in 1:length(groups)){

@@

The panels of figure \ref{figure1} illustrate the most common styles of ROC curve plots, and the associated design elements. I favor the use of gridlines and a diagonal reference line to facilitate accurate readings off of the axes. Direct labels are preferred over legends because they omit the additional cognitive step of matching line types or colors to labels. My \pkg{plotROC} package additionally provides plotting of cutoff values, which are displayed interactively with the web-based output option, and direct labels for print use. Exact confidence regions for points on the ROC curve are optionally calculated and displayed. Additionally, I use axis scales adjusted to be denser near the margins 0 and 1. In medical applications, it is often necessary to have a very low FPR (less than 10%, for instance), therefore the smaller scales are useful for accurately determining values near the margins. The next section details the usage of the \pkg{plotROC} \proglang{R} package and these features.

```{r figure1, fig.width = 6, fig.height = 6, echo = FALSE, fig.cap = "Illustration of design choices in plotting ROC curves. Panel A shows a sparse ROC curve, with no design additions inside the plotting region. The plot results in more white space than anything else. It is difficult to accurately determine values without reference lines. Panel B shows a plot comparing 2 curves, with different line types and a legend. AUCs are also given in the legend. Panels B and C add gridlines, diagonal reference lines, and direct labels. \\label{figure1}"}
library(plotROC)

@@

# Usage of the Package

@@

I start by creating an example data set. The marker I generate is moderately accurate for predicting disease status.

```{r dataset, echo = -1}
library(plotROC)
D.ex <- rbinom(100, size = 1, prob = .5)
M.ex <- rnorm(100, mean = D.ex)
```

Next I use the `calculate_roc` function to compute the empirical ROC curve. The disease status need not be coded as 0/1, but if it is not, `calculate_roc` assumes (with a warning) that the lowest value in sort order signifies disease-free status. This returns a dataframe with three columns: the cutoff values, the TPF and the FPF.

```{r calc}
roc.estimate <- calculate_roc(M.ex, D.ex)
str(roc.estimate)
```

The `rocdata` is passed to the `ggroc` function with an optional label. This creates a ggplot object of the ROC curve using the \pkg{ggplot2} package [@ggplot2].

@@

The `myrocplot` object can be used to create an interactive plot and display it in the Rstudio viewer or default web browser by passing it to the `plot_interactive_roc` function. Give the function an optional path to an html file as an argument called `file` to save the interactive plot as a complete web page. A screen shot of an interactive plot is shown in figure \ref{interact}. Hovering over the display shows the cutoff value at the point nearest to the cursor. Clicking makes the cutoff label stick until the next click, and if confidence regions are available, clicks will also display those as grey rectangles.

@@

An interactive ROC plot can be exported by using the `export_interactive_roc` function, which returns a character string containing the necessary \proglang{HTML} and \proglang{JavaScript}. The character string can be copy-pasted into an html document, or better yet, incorporated directly into a dynamic document using \pkg{knitr} [@knitr]. In a \pkg{knitr} document, it is necessary to use the `cat` function on the results and use the chunk options `results = 'asis'` and `fig.keep = 'none'` so that the interactive plot is displayed correctly. For documents that contain multiple interactive plots, it is necessary to assign each plot a unique name using the `prefix` argument of `export_interactive_roc`. This is necessary to ensure that the \proglang{JavaScript} code manipulates the correct svg elements. For examples of interactive plots and how to incorporate them into \pkg{knitr} documents, see the package vignette (`vignette("examples", package = "plotROC")`) or the web page https://sachsmc.github.io/plotROC/. The next code block shows an example \pkg{knitr} chunk that can be used in an .Rmd document to display an interactive plot.

```{r int-no, results = 'asis', fig.keep = 'none', eval = FALSE, verbatim = TRUE}
cat(
  export_interactive_roc(single.rocplot, 
                        prefix = "a")
  )
```

The same `ggroc` object that I called `myrocplot` can be used to generate an ROC plot suitable for use in print. It annotates the cutoff values and is completely in black and white. A simple example with the default options is shown in figure \ref{first}.

@@

If you have multiple tests of different types measured on the same subjects, you can use the `calculate_multi_roc` function to compute the empirical ROC curve for each test. It returns a list of data frames with the estimates and cutoff values. Then the `multi_ggroc` function creates the appropriate type of `ggplot` object. Confidence regions are not supported for multiple curves at the time of writing.

@@

Labels can be added easily with the `label` option of `multi_ggroc`. The length of the label element should match the number of plotted curves. The `multi_ggroc` object can be passed to the `plot_journal_roc` and the `export_interactive_roc` functions, as desired. The resulting plot is shown in figure \ref{multi}.

@@

Both `plot_journal_roc` and `export_interactive_roc` support a number of options to customize the look of the plots. By default, multiple curves are distinguished by different line types and direct labels. For multiple ROC curves that are similar to one another, those defaults can make it difficult to interpret the plots. Therefore, we also support colors and legends. The x- and y-axis can be changed by passing options to `ggroc` or `multi_ggroc`. The next code block illustrates the available options.

@@

I use the `ci = TRUE` option in `calculcate_roc` and `ggroc` to compute confidence regions for points on the ROC curve using the @clopper1934use exact method. Briefly, exact confidence intervals are calculated for the $FPF$ and $TPF$ separately, each at level $1 - \sqrt{1 - \alpha}$. Based on result 2.4 from @pepe2003statistical, the cross-product of these intervals yields a $100 * (1 - \alpha)$ percent rectangular confidence region for the pair. The significance level can be specified using the `alpha` option.

@@

For use in print, I pass a small vector of cutoff locations at which to display the confidence regions. This is shown in figure \ref{conf}.

@@

\pkg{plotROC} uses the \pkg{ggplot2} package to create the objects to be plotted. Therefore, themes and annotations can be added in the usual \pkg{ggplot2} way. A `plot_journal_roc` figure with a new theme, title, axis label, and AUC annotation is shown in figure \ref{annotate}.

@@

By default `calculate_roc` computes the empirical ROC curve. There are other estimation methods out there, as I have summarized in the introduction.  Any estimation method can be used, as long as the cutoff, the TPF and the FPF are returned. Then you can simply pass those values in a data frame to the `ggroc` function. For example, let us use the binormal method to create a smooth curve. This approach assumes that the test distribution is normal conditional on disease status.

@@

Then I can pass this data.frame to the `ggroc` function as before. The example is shown in figure \ref{binorm}.

@@

Another potential use of this approach is for plotting time-dependent ROC curves for time-to-event outcomes estimated as desribed in [@heagerty2000time].

# How it Works

\pkg{plotROC} makes use of \pkg{ggplot2} [@ggplot2], \pkg{gridSVG} [@gridsvg], and \pkg{d3.js} [@bostock2011d3] to create interactive plots. The first step in the process is to create `ggplot` objects using the `ggroc` or `multi_ggroc` functions. These functions return standard `ggplot` objects that include basic styling, hidden cutoff labels, and hidden confidence regions. They can be plotted and inspected in the \proglang{R} console. These form the basis for both the print versions and the interactive versions of the plots. Creating a print version by using the `plot_journal_roc` function simply makes visible a subset of the hidden cutoff labels and confidence regions, if available.

\pkg{plotROC} makes interactive plots by first converting the `ggplot` object into a scalable vector graphic (svg) object with the `grid.export` function. This function maps each element of the plot to a corresponding element of the svg markup language. I keep track of the names of the points and labels elements so that I can add interactivity using \pkg{d3.js} and \proglang{JavaScript}. The main interactive feature I wanted was to be able to display the cutoff labels at the points on the ROC curve closest to the mouse cursor.

There are many ways to solve this with \pkg{d3.js}, but I decided to use Voronoi polygons to map the cursor location to the nearest point on the ROC curve. The idea is that for the set of cutoff points along the ROC curve, the `d3.geom.voronoi` function chain computes a set of polygons overlaying the plotting region such that the area of each polygon contains the region of the plot closest to it's corresponding cutoff point. Hover events are bound to the polygons so that when the mouse cursor moves around the plotting region, the closest point on the ROC curve is made visible. Similarly, click events are bound to the polygons so that the appropriate confidence region is made visible upon clicking. The svg code and all necessary \proglang{JavaScript} code is returned in the character string provided by `export_interactive_roc`. Figure \ref{flow} outlines the \pkg{plotROC} process.

@@

This approach is similar to what is done in the \pkg{gridSVG} `grid.animate` function, which uses the svg `<animate />` tags. However, the available features were not sufficient for my needs, which is why I used \pkg{d3.js}. There are several other \proglang{R} packages that aim to create interactive figures. The authors of \pkg{animint} [@animint] created an extensive \proglang{JavaScript} library that creates plots in a similar way as \pkg{ggplot2}. A set of interactive features can be added to plots using \pkg{d3.js}. \pkg{ggvis} [@ggvis], \pkg{rCharts} [@rcharts], and the more recently released \pkg{htmlwidgets} [@htmlwidgets] all leverage existing charting libraries written in \proglang{JavaScript}. \pkg{qtlcharts} [@qtlcharts] uses a set of custom \proglang{JavaScript} and \pkg{d3.js} functions to visualize data from genetic experiments. Their general approach is to manipulate the data and create options in \proglang{R}, and then let the charting libraries or functions handle the rendering and interactivity. \pkg{plotROC} lets \proglang{R} do the rendering, allowing the figures to be consistent across print and web-based media, and retaining the distinctive \proglang{R} style. This also allows users to manipulate the figures directly in \proglang{R} to suit their needs, using tools that are more accessible and familiar to most \proglang{R} users.

@@

## Reproducibility Note

This manuscript is completely reproducible using the source files. The output below indicates the \proglang{R} packages and versions used. Compiling the pdf output also requires `pandoc` version 1.13.1 and `pdflatex`.

>>>>

---
title: '\pkg{plotROC}: A Better Tool for Plotting ROC Curves'
shorttitle: '\pkg{plotROC}: A Better Tool for Plotting ROC Curves'
plaintitle: 'plotROC: A Better Tool for Plotting ROC Curves'
author:
- name: Michael C. Sachs
  affil: Biometric Research Branch, Division of Cancer Treatment and Diagnosis, National Cancer Institute
address:
- Michael C. Sachs
- 9609 Medical Center Drive, MSC 9735
- Bethesda, MD 20892
- "telephone: 240-276-7888"
email: michael.sachs@nih.gov
abstract: 'Plots of the receiver operating characteristic (ROC) curve are ubiquitous in medical research. Designed to simultaneously display the operating characteristics at every possible value of a continuous diagnostic test, ROC curves are used in oncology to evaluate screening, diagnostic, prognostic and predictive biomarkers. I reviewed a sample of ROC curve plots from the major oncology journals in order to assess current trends in usage and design elements. My review suggests that ROC curve plots are often ineffective as statistical charts and that poor design obscures the relevant information the chart is intended to display. I describe my new \proglang{R} package that was created to address the shortcomings of existing tools. The package has functions to create informative ROC curve plots, with sensible defaults, for use in print or as an interactive web-based plot. A web application was developed to reach a broader audience of scientists who do not use \proglang{R}.'
keywords: ROC curves; graphics; interactive; plots
bibliography: plotroc
...

```{r setup, echo = FALSE, include = FALSE}
library(knitr)
library(stringr)

knit_hooks$set(plot = function(x, options) {
  if ('tikz' %in% options$dev && !options$external) {
    hook_plot_tex(x, options)
  } else hook_plot_md(x, options)
})

options(prompt = "R> ")

opts_chunk$set(dev="tikz", prompt = TRUE)

render_sweave()

@@

## About ROC curves

@@

## Design considerations

@@

## Existing plotting software

The ROC curve plot is, at the most basic level, a line graph. Therefore, once the appropriate statistics are estimated, existing plotting functions can be used to create an ROC curve plot. Viewers can identify ROC plots through context, by observing the shape of the line, and through the addition of axis labels, titles, legends, and so on. In my review of the oncology literature, I observed plots with the distinctive characteristics of the plotting functions from Microsoft Excel [@excel], \proglang{SAS} [@sas], \proglang{SPSS} [@spss], and the base \proglang{R} plotting functions [@arr].

There are several \proglang{R} packages related to ROC curve estimation that contain dedicated plotting functions. The \pkg{ROCR} package [@rocr] plots the FPF _versus_ TPF, as usual, and then takes the interesting approach of encoding the cutoff values as a separate color scale along the ROC curve itself. A legend for the color scale is placed along the vertical axis on the right of the plotting region. Due to the popularity of \pkg{ROCR}, my package supports the performance objects as returned by \pkg{ROCR}. The \pkg{pROC} package [@pROC] is mainly focused on estimating confidence intervals and regions for restricted ranges of the ROC curve. The plotting methods therein use the base \proglang{R} plotting functions to create nice displays of the curves along with shaded confidence regions. My \pkg{plotROC} package uses the \pkg{ggplot2} [@ggplot2] plotting library to create clear, informative ROC plots, with interactive features for use on the web, and sensible defaults for use in print.

@@

Each image consisted of a single ROC curve plot or a panel of multiple plots. Each plot was inspected manually for the following design features: the number of curves displayed, the type of axis labels (sensitivity/ 1 - specificity or true/false positive fractions), presence or absence of grid lines, presence or absence of a diagonal guide line, whether any cutpoints were indicated, the type of curve label (legend or direct label), and presence of other textual annotations such as the AUC. The numerical results of the survey are summarized in Table \ref{table1}.

@@

    if(tab) groupsa <- paste0("\t", groups) else groupsa <- groups
    blah <- data.frame(lab = groupsa, `Percent (count)`  = NA, check.names = FALSE, stringsAsFactors = FALSE)
    N <- sum(!is.na(var))
    for(k in 1:length(groups)){

@@

The panels of Figure \ref{figure1} illustrate the most common styles of ROC curve plots, and the associated design elements. I favor the use of gridlines and a diagonal reference line to facilitate accurate readings off of the axes. Direct labels are preferred over legends because they omit the additional cognitive step of matching line types or colors to labels. My \pkg{plotROC} package additionally provides plotting of cutoff values, which are displayed interactively with the web-based output option, and direct labels for print use. Exact confidence regions for points on the ROC curve are optionally calculated and displayed. Additionally, I use axis scales adjusted to be denser near the margins 0 and 1. In medical applications, it is often necessary to have a very low FPR (less than 10%, for instance), therefore the smaller scales are useful for accurately determining values near the margins. The next section details the usage of the \pkg{plotROC} \proglang{R} package and these features.

```{r figure1, fig.width = 6, fig.height = 6, echo = FALSE, fig.cap = "Illustration of design choices in plotting ROC curves. Panel A shows a sparse ROC curve, with no design additions inside the plotting region. The plot results in more white space than anything else. It is difficult to accurately determine values without reference lines. Panel B shows a plot comparing 2 curves, with different line types and a legend. AUCs are also given in the legend. Panels C and D add gridlines, diagonal reference lines, and direct labels. \\label{figure1}"}
library(plotROC)

@@

# Usage of the package

@@

I start by creating an example data set. The marker I generate is moderately accurate for predicting disease status. Next I use the \code{calculate_roc} function to compute the empirical ROC curve. The disease status need not be coded as 0/1, but if it is not, \code{calculate_roc} assumes (with a warning) that the lowest value in sort order signifies disease-free status. This returns a dataframe with three columns: the cutoff values, the TPF and the FPF.

```{r dataset, echo = -1}
library(plotROC)
D.ex <- rbinom(100, size = 1, prob = .5)
M.ex <- rnorm(100, mean = D.ex)
roc.estimate <- calculate_roc(M.ex, D.ex)
head(roc.estimate)
```

The \code{roc.estimate} object is passed to the \code{ggroc} function with an optional label. This creates a \code{ggplot} object of the ROC curve using the \pkg{ggplot2} package [@ggplot2].

@@

The \code{single.rocplot} object can be used to create an interactive plot and display it in the Rstudio viewer or default web browser by passing it to the \code{plot_interactive_roc} function. Give the function an optional path to an html file as an argument called \code{file} to save the interactive plot as a complete web page. A screen shot of an interactive plot is shown in Figure \ref{interact}. Hovering over the display shows the cutoff value at the point nearest to the cursor. Clicking makes the cutoff label stick until the next click, and if confidence regions are available, clicks will also display those as grey rectangles.

@@

An interactive ROC plot can be exported by using the \code{export_interactive_roc} function, which returns a character string containing the necessary \proglang{HTML} and \proglang{JavaScript}. The character string can be copy-pasted into an html document, or better yet, incorporated directly into a dynamic document using \pkg{knitr} [@knitr]. In a \pkg{knitr} document, it is necessary to use the \code{cat} function on the results and use the chunk options \code{results = 'asis'} and \code{fig.keep = 'none'} so that the interactive plot is displayed correctly. For documents that contain multiple interactive plots, it is necessary to assign each plot a unique name using the \code{prefix} argument of \code{export_interactive_roc}. This is necessary to ensure that the \proglang{JavaScript} code manipulates the correct svg elements. For examples of interactive plots and how to incorporate them into \pkg{knitr} documents, see the package vignette (\code{vignette("examples", package = "plotROC")}) or the web page https://sachsmc.github.io/plotROC/. The next code block shows an example \pkg{knitr} chunk that can be used in an .Rmd document to display an interactive plot.

\begin{Code}
`r paste("\x60\x60\x60{r int-no, results = 'asis', fig.keep = 'none'}")`
cat(
  export_interactive_roc(single.rocplot, 
                        prefix = "a")
  )
`r paste("\x60\x60\x60")`
\end{Code}

The same \code{ggroc} object that I called \code{single.rocplot} can be used to generate an ROC plot suitable for use in print. It annotates the cutoff values and is completely in black and white. A simple example with the default options is shown in Figure \ref{first}.

@@

If you have multiple tests of different types measured on the same subjects, you can use the \code{calculate_multi_roc} function to compute the empirical ROC curve for each test. It returns a list of data frames with the estimates and cutoff values. Then the \code{multi_ggroc} function creates the appropriate type of \code{ggplot} object. Confidence regions are not supported for multiple curves at the time of writing.

@@

Labels can be added easily with the \code{label} option of \code{multi_ggroc}. The length of the label element should match the number of plotted curves. The \code{multi_ggroc} object can be passed to the \code{plot_journal_roc} and the \code{export_interactive_roc} functions, as desired. The resulting plot is shown in Figure \ref{multi}.

@@

Both \code{plot_journal_roc} and \code{export_interactive_roc} support a number of options to customize the look of the plots. By default, multiple curves are distinguished by different line types and direct labels. For multiple ROC curves that are similar to one another, those defaults can make it difficult to interpret the plots. Therefore, we also support colors and legends. The x- and y-axis can be changed by passing options to \code{ggroc} or \code{multi_ggroc}. The next code block illustrates the available options.

@@

I use the \code{ci = TRUE} option in \code{calculcate_roc} and \code{ggroc} to compute confidence regions for points on the ROC curve using the @clopper1934use exact method. Briefly, exact confidence intervals are calculated for the $FPF$ and $TPF$ separately, each at level $1 - \sqrt{1 - \alpha}$. Based on result 2.4 from @pepe2003statistical, the cross-product of these intervals yields a $100 * (1 - \alpha)$ percent rectangular confidence region for the pair. The significance level can be specified using the \code{alpha} option.

@@

For use in print, I pass a small vector of cutoff locations at which to display the confidence regions. This is shown in Figure \ref{conf}.

@@

\pkg{plotROC} uses the \pkg{ggplot2} package to create the objects to be plotted. Therefore, themes and annotations can be added in the usual \pkg{ggplot2} way. A \code{plot_journal_roc} figure with a new theme, title, axis label, and AUC annotation is shown in Figure \ref{annotate}.

@@

By default \code{calculate_roc} computes the empirical ROC curve. There are other estimation methods out there, as I have summarized in the introduction.  Any estimation method can be used, as long as the cutoff, the TPF and the FPF are returned. Then you can simply pass those values in a data frame to the \code{ggroc} function. For example, let us use the binormal method to create a smooth curve. This approach assumes that the test distribution is normal conditional on disease status.

@@

Then I can pass this \code{data.frame} to the \code{ggroc} function as before. The example is shown in Figure \ref{binorm}.

@@

\pkg{plotROC} also supports the use of the \pkg{ROCR} estimation method. Objects as returned by the function \code{rocr::performance} can be passed directly to \code{ggroc}:

```{r rocr, message = FALSE, warning = FALSE}
library(ROCR)
rocr.est <- performance(prediction(M.ex, D.ex), "tpr", "fpr")
rocr.plot <- ggroc(rocr.est, label = "ROCR object")
```

Another potential use of this approach is for plotting time-dependent ROC curves for time-to-event outcomes estimated as desribed in [@heagerty2000time]. Here is an example using the \pkg{survivalROC} package [@survroc] for estimation:

```{r survival}
library(survivalROC)
survT <- rexp(350, 1/5)
cens <- rbinom(350, 1, .1)
M <- -8 * sqrt(survT) + rnorm(350, sd = survT)
sroc <- lapply(c(2, 5, 10), function(t){ 
  stroc <- survivalROC(Stime = survT, status = cens, marker = M, 
                       predict.time = t, method = "NNE", 
                       span = .25 * 350^(-.2))
  data.frame(TPF = stroc[["TP"]], FPF = stroc[["FP"]], 
             c = stroc[["cut.values"]], 
             time = rep(stroc[["predict.time"]], 
                        length(stroc[["FP"]])))
  })

surv.plot <- multi_ggroc(sroc, label = paste("t =", c(2, 5, 10)))
```

# How it works

\pkg{plotROC} makes use of \pkg{ggplot2} [@ggplot2], \pkg{gridSVG} [@gridsvg], and \pkg{d3.js} [@bostock2011d3] to create interactive plots. The first step in the process is to create \code{ggplot} objects using the \code{ggroc} or \code{multi_ggroc} functions. These functions return standard \code{ggplot} objects that include basic styling, hidden cutoff labels, and hidden confidence regions. They can be plotted and inspected in the \proglang{R} console. These form the basis for both the print versions and the interactive versions of the plots. Creating a print version by using the \code{plot_journal_roc} function simply makes visible a subset of the hidden cutoff labels and confidence regions, if available.

\pkg{plotROC} makes interactive plots by first converting the \code{ggplot} object into a scalable vector graphic (svg) object with the \code{gridSVG::grid.export} function. This function maps each element of the plot to a corresponding element of the svg markup language. I keep track of the names of the points and labels elements so that I can add interactivity using \pkg{d3.js} and \proglang{JavaScript}. The main interactive feature I wanted was to be able to display the cutoff labels at the points on the ROC curve closest to the mouse cursor.

There are many ways to solve this with \pkg{d3.js}, but I decided to use Voronoi polygons to map the cursor location to the nearest point on the ROC curve. The idea is that for the set of cutoff points along the ROC curve, the \code{d3.geom.voronoi} function chain computes a set of polygons overlaying the plotting region such that the area of each polygon contains the region of the plot closest to it's corresponding cutoff point. Hover events are bound to the polygons so that when the mouse cursor moves around the plotting region, the closest point on the ROC curve is made visible. Similarly, click events are bound to the polygons so that the appropriate confidence region is made visible upon clicking. The svg code and all necessary \proglang{JavaScript} code is returned in the character string provided by \code{export_interactive_roc}. Figure \ref{flow} outlines the \pkg{plotROC} process.

@@

This approach is similar to what is done in the \pkg{gridSVG} \code{grid.animate} function, which uses the svg \code{<animate />} tags. However, the available features were not sufficient for my needs, which is why I used \pkg{d3.js}. There are several other \proglang{R} packages that aim to create interactive figures. The authors of \pkg{animint} [@animint] created an extensive \proglang{JavaScript} library that creates plots in a similar way as \pkg{ggplot2}. A set of interactive features can be added to plots using \pkg{d3.js}. \pkg{ggvis} [@ggvis], \pkg{rCharts} [@rcharts], and the more recently released \pkg{htmlwidgets} [@htmlwidgets] all leverage existing charting libraries written in \proglang{JavaScript}. \pkg{qtlcharts} [@qtlcharts] uses a set of custom \proglang{JavaScript} and \pkg{d3.js} functions to visualize data from genetic experiments. Their general approach is to manipulate the data and create options in \proglang{R}, and then let the charting libraries or functions handle the rendering and interactivity. \pkg{plotROC} lets \proglang{R} do the rendering, allowing the figures to be consistent across print and web-based media, and retaining the distinctive \proglang{R} style. This also allows users to manipulate the figures directly in \proglang{R} to suit their needs, using tools that are more accessible and familiar to most \proglang{R} users.

@@

## Reproducibility note

This manuscript is completely reproducible using the source files. The output below indicates the \proglang{R} packages and versions used. Compiling the pdf output also requires \pkg{pandoc} version 1.13.1 and \pkg{pdflatex}.
